---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# phsmethods

[![GitHub release (latest by date)](https://img.shields.io/github/v/release/Public-Health-Scotland/phsmethods)](https://github.com/Public-Health-Scotland/phsmethods/releases/latest)
[![Build Status](https://github.com/Public-Health-Scotland/phsmethods/workflows/R-CMD-check/badge.svg)](https://github.com/Public-Health-Scotland/phsmethods/actions)
[![codecov](https://codecov.io/gh/Public-Health-Scotland/phsmethods/branch/master/graph/badge.svg)](https://codecov.io/gh/Public-Health-Scotland/phsmethods)

`phsmethods` contains functions for commonly undertaken analytical tasks in [Public Health Scotland (PHS)](https://www.publichealthscotland.scot/):

- `create_age_groups()` categorises ages into groups
- `chi_check()` assesses the validity of a CHI number
- `chi_pad()` adds a leading zero to nine-digit CHI numbers
- `sex_from_chi()` extracts the sex of a person from a CHI number
- `file_size()` returns the names and sizes of files in a directory
- `extract_fin_year()` assigns a date to a financial year in the format `YYYY/YY`
- `match_area()` converts geography codes into area names
- `format_postcode()` formats improperly recorded postcodes
- `qtr()`, `qtr_end()`, `qtr_next()` and `qtr_prev()` assign a date to a quarter
- `age_calculate` calculates age between two dates
- `dob_from_chi` extracts Date of Birth (DoB) from the CHI number
- `age_from_chi` extracts age from the CHI number

`phsmethods` can be used on both the [server](https://rstudio.nhsnss.scot.nhs.uk/) and desktop versions of RStudio.

## Installation

To install `phsmethods`, the package `remotes` is required, and can be installed with `install.packages("remotes")`.

You can then install `phsmethods` on RStudio server from GitHub with:

```{r gh-installation, eval = FALSE}
remotes::install_github("Public-Health-Scotland/phsmethods",
  upgrade = "never"
)
```

Network security settings may prevent `remotes::install_github()` from working on RStudio desktop. If this is the case, `phsmethods` can be installed by downloading the [zip of the repository](https://github.com/Public-Health-Scotland/phsmethods/archive/master.zip) and running the following code (replacing the section marked `<>`, including the arrows themselves):

```{r source-installation, eval = FALSE}
remotes::install_local("<FILEPATH OF ZIPPED FILE>/phsmethods-master.zip",
  upgrade = "never"
)
```

## Using phsmethods

Load `phsmethods` using `library()`:

```{r load}
library(phsmethods)
```

To access the help file for any of `phsmethods`' functions, type `?function_name` into the RStudio console after loading the package:

```{r help, eval = FALSE}
?extract_fin_year
?format_postcode
```

### create_age_groups

```{r create_age_groups}
ages <- c(54, 7, 77, 1, 26, 101)

# By default create_age_groups goes in 5 year increments from 0 to 90+
create_age_groups(ages)

# But these settings can be changed
create_age_groups(ages, from = 0, to = 80, by = 10)
```

### chi_check

```{r chi_check, message = FALSE, warning = FALSE}
# chi_check returns specific feedback on why a CHI number might be invalid
library(dplyr)
chi_data <- tibble(chi = c("0101011237", "0101336489", "3213201234", "123456789", "12345678900", "010120123?", NA))
chi_data %>% 
  mutate(validity = chi_check(chi))
```

### chi_pad

```{r chi_pad}
chi_numbers <- c("101011237", "101201234", "123223", "abcdefghi", "12345tuvw")
# Only nine-digit characters comprised exclusively of numeric digits are prefixed with a zero
chi_pad(chi_numbers)
```


### sex_from_chi

```{r sex_from_chi}
# By default it will check that the CHI is valid before extracting the sex.
library(dplyr)
chi_data <- tibble(chi = c("0101011237", "0101336489", "123456789", "12345678900", "010120123?", NA))

chi_data %>% 
  mutate(chi_sex = sex_from_chi(chi))

# Use custom values for male and female
chi_data %>% 
  mutate(chi_sex = sex_from_chi(chi, male_value = "M", female_value = "F"))

# Alternatively return the result as a factor (with labels 'Male' and 'Female')
chi_data %>% 
  mutate(chi_sex = sex_from_chi(chi, as_factor = TRUE))
```


### file_size

```{r file_size}
# Names and sizes of all files in the tests/testthat/files folder
file_size(testthat::test_path("files"))

# Names and sizes of Excel files only in the tests/testthat/files folder
file_size(testthat::test_path("files"), pattern = "\\.xlsx?$")
```

### extract_fin_year

```{r extract_fin_year}
dates <- lubridate::dmy(c(21012017, 04042017, 17112017))
extract_fin_year(dates)
```

### match_area

```{r match_area}
match_area("S13002781")

d <- tibble(code = c("S02000656", "S02001042", "S08000020", "S12000013", "S13002605"))
d %>% 
  mutate(name = match_area(code))
```

### format_postcode

```{r format_postcode, warning = FALSE}
# The default is pc7 format
format_postcode("G26QE")

# But pc8 format can also be applied
format_postcode(c("KA89NB", "PA152TY"), format = "pc8")

# postcode accounts for irregular spacing and lower case letters
e <- tibble(pc = c("G 4 2 9 B A", "g207al", "Dg98bS", "DD37J    y"))
e %>% 
  mutate(pc = format_postcode(pc))
```

### qtr, qtr_end, qtr_next and qtr_prev
```{r qtr}
f <- lubridate::dmy(c(26032012, 04052012, 23092012))

# qtr returns the current quarter and year
# The default is long format
qtr(f)

# But short format can also be applied
qtr(f, format = "short")


# qtr_end returns the last month in the quarter
qtr_end(f)
qtr_end(f, format = "short")


# qtr_next returns the next quarter
qtr_next(f)
qtr_next(f, format = "short")


# qtr_prev returns the previous quarter
qtr_prev(f)
qtr_prev(f, format = "short")
```

### age_calculate
```{r age_calculate}
birth_date <- lubridate::ymd("2020-02-29")
end_date <- lubridate::ymd("2022-02-21")

# Change the argument of date_class can make a difference.
age_calculate(birth_date, end_date, round_down = FALSE, date_class = "period") * 365.25
age_calculate(birth_date, end_date, round_down = FALSE, date_class = "duration") * 365.25

# If the start day is leap day (February 29th), age increases on 1st March every year. 
leap1 <- lubridate::ymd("2020-02-29")
leap2 <- lubridate::ymd("2022-02-28")

age_calculate(leap1, leap2, date_class = "period")
```

### dob_from_chi
```{r dob_from_chi}
dob_from_chi("0101336489")

library(tibble)
library(dplyr)
data <- tibble(chi = c(
 "0101336489",
 "0101405073",
 "0101625707"
), adm_date = as.Date(c(
 "1950-01-01",
 "2000-01-01",
 "2020-01-01"
)))

data %>%
 mutate(chi_dob = dob_from_chi(chi))

data %>%
 mutate(chi_dob = dob_from_chi(chi,
   min_date = as.Date("1930-01-01"),
   max_date = adm_date
 ))
```

### age_from_chi
```{r age_from_chi}
age_from_chi("0101336489")

library(tibble)
library(dplyr)
data <- tibble(chi = c(
 "0101336489",
 "0101405073",
 "0101625707"
), dis_date = as.Date(c(
 "1950-01-01",
 "2000-01-01",
 "2020-01-01"
)))

data %>%
 mutate(chi_age = age_from_chi(chi))

data %>%
 mutate(chi_age = age_from_chi(chi, min_age = 18, max_age = 65))

data %>%
 mutate(chi_age = age_from_chi(chi,
   ref_date = dis_date
 ))
```

## Contributing to phsmethods

At present, the maintainer of this package is [Tina Fu](https://github.com/Tina815).

This package is intended to be in continuous development and contributions may be made by anyone within PHS. If you would like to contribute, please first create an [issue](https://github.com/Public-Health-Scotland/phsmethods/issues) on GitHub and assign **both** of the package maintainers to it. This is to ensure that no duplication of effort occurs in the case of multiple people having the same idea. The package maintainers will discuss the issue and get back to you as soon as possible.

While the most obvious and eyecatching (as well as intimidating) way of contributing is by writing a function, this isn't the only way to make a useful contribution. Fixing typos in documentation, for example, isn't the most glamorous way to contribute, but is of great help to the package maintainers. Please see this [blog post by Jim Hester](https://www.tidyverse.org/blog/2017/08/contributing/) for more information on getting started with contributing to open-source software.

When contributing, please create a [branch](https://github.com/Public-Health-Scotland/phsmethods/branches) in this repository and carry out all work on it. Please ensure you have linked RStudio to your GitHub account using `usethis::edit_git_config()` prior to making your contribution. When you are ready for a review, please create a [pull request](https://github.com/Public-Health-Scotland/phsmethods/pulls) and assign **both** of the package maintainers as reviewers. One or both of them will conduct a review, provide feedback and, if necessary, request changes before merging your branch.

Please be mindful of information governance when contributing to this package. No data files (aside from publicly available and downloadable datasets or unless explicitly approved), server connection details, passwords or person identifiable or otherwise confidential information should be included anywhere within this package or any other repository (whether public or private) used within PHS. This includes within code and code commentary. For more information on security when using git and GitHub, and on using git and GitHub for version control more generally, please see the [Transforming Publishing Programme](https://www.isdscotland.org/Products-and-Services/Transforming-Publishing-Programme/)'s [Git guide](https://Public-Health-Scotland.github.io/git-guide/) and [GitHub guidance](https://github.com/Public-Health-Scotland/GitHub-guidance).

Please feel free to add yourself to the 'Authors' section of the `Description` file when contributing. As a rule of thumb, please assign your role as author (`"aut"`) when writing an exported function, and as contributor (`"ctb"`) for anything else.

`phsmethods` will, as much as possible, adhere to the [tidyverse style guide](https://style.tidyverse.org/) and the [rOpenSci package development guide](https://devguide.ropensci.org/). The most pertinent points to take from these are:

- All function names should be in lower case, with words separated by an underscore
- Put a space after a comma, never before
- Put a space before and after infix operators such as `<-`, `==` and `+`
- Limit code to 80 characters per line
- Function documentation should be generated using [`roxygen2`](https://github.com/r-lib/roxygen2)
- All functions should be tested using [`testthat`](https://github.com/r-lib/testthat)
- The package should always pass `devtools::check()`

It's not necessary to have experience with GitHub or of building an R package to contribute to `phsmethods`. If you wish to contribute code then, as long as you can write an R function, the package maintainers can assist with error handling, writing documentation, testing and other aspects of package development. It is advised, however, to consult [Hadley Wickham's R Packages book](https://r-pkgs.org/) prior to making a contribution. It may also be useful to consult the [documentation](https://github.com/Public-Health-Scotland/phsmethods/tree/master/R) and [tests](https://github.com/Public-Health-Scotland/phsmethods/tree/master/tests/testthat) of existing functions within this package as a point of reference.

Please note that this README may fail to 'Knit' at times as a result of network security settings. This will likely be due to the badges for the package's release version, continuous integration status and test coverage at the top of the document. You should only make edits to the `.Rmd` version, you can then knit yourself, or a GitHub action should run and knit it for you when you open a pull request.

